
We will first prove Proposition~\ref{proposition:np}.

\begin{proof}[of Proposition~\ref{proposition:np}]
We will prove the hardness from \tmatch: an \np-hard problem where we are given
a universe $U$ and family $\mathcal{M}$ of sets of size $3$, and we are asked
to find a maximum disjoint cover. 

Assume that we are given an instance $(U, \mathcal{M})$ of \tmatch.  Let $n =
\abs{U}$ and $m = \abs{\mathcal{M}}$. We can safely assume that $n$ is divisible by 3.

Define $h = {n \choose 3}$ the number of possible
sets of size $3$ over $U$. Let $\ell = {n - 1 \choose 2}$ be the number of possible
sets of size $3$ containing a fixed vertex in $U$.
Let us define $t = 8m$ and $s = 2t$.

\emph{Graph construction:}
The graph constists of two major parts.

The first part entails 7 vertex sets: we start with $A_1$ and $A_2$ with $\abs{A_1} = 3$ and $\abs{A_2}
= \ell - 1$.
A single vertex, say $v$, 
is connected to an additional vertex, say $w$. We write $A_3 = \set{v}$ and $A_4 = \set{w}$.
Futhermore, $w$ is connected to a biclique of $A_5 = K(s, s)$.
%Each vertex $a \in A_1$ is connected to its own clique of size $h$; the vertices in these cliques form $A_6$.

Each vertex $u \in A_1$ and $v \in A_2$ is connected with a \emph{fat} edge:
$t$ copies of a path $u$--$x$--$v$, where $x$ is a vertex unique to the path.
These vertices form $A_6$.
Similarly, each vertex $u \in A_1$ and $v \in A_3$ is connected with a \emph{fat} edge;
the intermediate vertices form $A_7$.

%Finally, one
%vertex, say $b^* \in B$ is connected to an additional vertex, who is also connected
%to a biclique $K(s, s)$.

The final graph consists of $t$ copies of the first part. We redefine $A_i$ to be
the union of the corresponding groups in each copy.

The second part is very similar to the first.  It entails the vertex sets
$B_1$, $B_2$, and $B_3$ with $\abs{B_1} = n$, $\abs{B_2} = h - m$ $\abs{B_3} = m$.
Here $B_1$ corresponds to the universe $U$, and $B_2$ and $B_3$ correspond to
triplets of elements in $U$. Each vertex in $B_3$ is connected to its own vertex.
These vertices form $B_4$. Each vertex in $B_4$ is also connected to its own dedicated biclique
$K(s, s)$. The vertices in bicliques form $B_5$.
A vertex in $B_2$ is connected to the corresponding vertices in $B_1$ with a fat edge
of size $t$. The intermediate vertices form $B_6$.
Similarly, a vertex in $B_3$ is connected to the corresponding vertices in $B_1$ with a fat edge
of size $t$. The intermediate vertices form $B_7$.



\iffalse
The first part entails vertex sets $A$ and $B$ with $\abs{A} = 3$ and $\abs{B}
= \ell$.  Each vertex $a \in A$ and $b \in B$ is connected with a \emph{fat} edge:
$t$ copies of a path $a$--$x$--$b$, where $x$ is a vertex unique to the path.
Each vertex $a \in A$ is connected to its own clique of size $h$.  Finally, one
vertex, say $b^* \in B$ is connected to an additional vertex, who is also connected
to a biclique $K(s, s)$.

The final graph consists of $t$ copies of the first part.

The second part is very similar to the first.  It entails the vertex sets $A'$
and $B'$ with $\abs{A'} = n$ and $\abs{B'} = h$. Here $A'$ corresponds to the
universe $U$.  Each vertex $a \in A'$ and $b \in B'$ is connected with a
\emph{fat} edge of size $t$.  Finally, let $C'$ be the subset of $B'$ that
corresponds to the sets in $\mathcal{M}$.  Each vertex $c \in C'$ is connected
to its own unique vertex. We will denote these vertices by $O'$.
Each vertex in $O'$ is also connected to its own dedicated bi-clique
$K(s, s)$. 
\fi

\emph{Role assignment:}
Let us now define a role assignment, later this will turn out to be the optimal
assignment. Let $\mathcal{W}$, not necessarily a subset of $\mathcal{M}$, be a
matching covering completely $U$.

Let $C_3$ be the vertices in $B_2 \cup B_3$ corresponding to $\mathcal{W}$.
Let $C_1 = B_1$, $C_2 = (B_2 \cup B_3) \setminus C_3$ $C_4 = B_4$, $C_5 = B_5$.
Let $C_6$ be the vertices in $B_6 \cup B_7$ adjacent to $C_3$, and set $C_7 = (B_6 \cup B_7) \setminus C_6$.
We define $\role(v) = i$, where $v \in A_i \cup C_i$.
Let us write $\role_{\mathcal{W}} = \role$.

Let us compute the cost of $\role$.
To that end, write $c_{ij}$ to be the cost incurred by the $j$th component of the $i$th role.
The only non-zero costs are $c_{24}$, $c_{42}$, $c_{34}$, and $c_{43}$.
Define $n_i = \abs{\set{v ; \role(v) = i}}$,
and let $\alpha = \abs{A_3 \cup (B_3 \cap C_3)}$ and $\beta = \abs{B_3 \cap C_2}$.
We can express the cost as
\[
	c_{24} + c_{34} + c_{42} + c_{43} = f(\beta, n_2) + f(\alpha, n_3) + 2f(\alpha, n_4),
	%c_{24} = f(\beta, n_2)
	%c_{34} = f(\alpha, n_3)
	%c_{42} = f(\alpha, n_4)
\]
where $f(x, y) = x(y - x)/y$.
The counts $n_i$ do not depend on $\mathcal{W}$.
Since $\alpha + \beta = n_4$, so the cost of depends on $\mathcal{W}$ only via $\alpha$.
We have $2\alpha \geq 2t \geq n_3, n_4$ and $2\beta \leq 2m \leq t \leq n_2$.
Since $f(x, y)$ is a parabola peaking at $x = y / 2$,
the cost decreases as $\alpha$ increases.

Since $f(x, y) \leq \min(x, y / 2 - x)$, we can upper-bound the cost
by $n_3 + 2n_4 - 3\alpha + \beta < 4m$.

\iffalse

Let $\alpha = \abs{A_3 \cup C_3}$ be the number of vertices with $\role(v) = 3$.
Let $\beta = \abs{A_2 \cup C_2}$ be the number of vertices with $\role(v) = 2$.
Define also $\alpha_1 = \abs{A_3 \cup (B_3 \cap C_3)}$,
and $\beta_1 = \abs{B_3 \cap C_2}$.
A laborous comptuation reveals that the cost of $\role$ is equal to
\begin{equation}
\label{eq:cost}
	f(\alpha_1, \alpha) + f(\beta_1, \beta) + 2f(\alpha_1, \alpha_1 + \beta_1),
\end{equation}
where $f(x, y) = 2x(y - x)/y$.
Note that $\alpha$, $\beta$, and $\alpha_1 + \beta_1 = \abs{A_2} + \abs{B_3}$ are all constant.
Moreover, since
$\alpha_1 \geq \alpha / 2$,
$\beta_1 \leq \beta / 2$,
$\alpha_1 \geq \beta_1$, Eq.~\ref{eq:cost} decreases as $\abs{B_3 \cap C_3}$ increases.

Since $f(x, y) \geq x$ for $x \leq y / 2$,
we can upper-bound the cost by
\[
\begin{split}
	\alpha - \alpha_1 + \beta_1 + 2\beta_1 & = \abs{A_3 \cup C_3} - \abs{A_3 \cup (B_3 \cap C_3)} + 
	t + n / 3 - (t + \abs{D}) + 3(m - \abs{D}) \\
	& = n / 3 + 3m - 4\abs{D} < 4m.
\end{split}
\]

\emph{Role assignment:}
Let us now define a role assignment, this will turn out to be optimal
assignment.  Let $\mathcal{W} \subseteq \mathcal{U}$ be the maximum matching.
If $\mathcal{W}$ does not cover completely $\mathcal{U}$, then let $\mathcal{W}'$
be additional sets outside $\mathcal{M}$ completing the matching.
Let $D$ be the vertices in $B'$ corresponding to $\mathcal{W}$.
Let $D'$ be the vertices in $B'$ corresponding to $\mathcal{W}'$.
Let $K$ be the union of all bi-cliques.
\[
	\role(v) =
\begin{cases}
1 & v \in K,\\
2 & v \text{ adjacent to } K,\\
3 & v \in A \text{ or } v \in A',\\
4 & v \in D \cup D', \text{ or } v = b^*, \\
5 & v \in B' \setminus (D \cup D'), \text{ or } v \in B \setminus \set{b^*},\\
6 & v \text{ is a fat edge adjacent to } w, \role(w) = 4,\\
7 & v \text{ is a fat edge adjacent to } w, \role(w) = 5.\\
\end{cases}
\]

Define the roles as follows: The roles of vertices in bi-cliques $K(s, s)$ are
set to 2, the roles of adjacent vertices to the bi-cliques are set to 1. The
roles of vertices in $A$ and $A'$ are set to 3. The roles of vertices
corresponding to $\mathcal{W}$, as well as $b^*$, are set to $4$.
The roles for the remaining vertices in $B$ and $B'$ are set to 

Let us compute the cost of $\role$.
Let $\alpha = t + n / 3$ be the number of vertices with $\role(v) = 4$.
Among these vertices,
let $\alpha_1 = t + \abs{D}$ be the number of vertices adjacent to a vertex with
a role of 2.
Let $\beta = t\ell + k - \alpha$ be the number of vertices with $\role(v) = 5$.
Among these vertices,
let $\beta_1 = m - \abs{D}$ be the number of vertices adjacent to a vertex with
a role of 2.
A laborous comptuation reveals that the cost of $\role$ is equal to
\begin{equation}
\label{eq:cost}
	f(\alpha_1, \alpha) + f(\beta_1, \beta) + 2f(\alpha_1, \alpha_1 + \beta_1),
\end{equation}
where $f(x, y) = 2x(y - x)/y$. Note that $\alpha$, $\beta$, and $\alpha_1 + \beta_1$ are all constant.
Moreover, since
$\alpha_1 \geq \alpha / 2$,
$\beta_1 \leq \alpha / 2$,
$\alpha_1 \geq \beta$, Eq.~\ref{eq:cost} decreases as $\abs{D}$ increases.


Since $f(x, y) \geq x$ for $x \leq y / 2$,
we can upper-bound the cost by
\[
\begin{split}
	\alpha - \alpha_1 + \beta_1 + 2\beta_1 & = t + n / 3 - (t + \abs{D}) + 3(m - \abs{D}) \\
	& = n / 3 + 3m - 4\abs{D} < 4m.
\end{split}
\]

\fi

\emph{Role $\role$ is optimal:}
Let $\role^*$ be the optimal role assignment with a cost of $\sigma$.  Consider
that if instead of selecting optimal centroids for $\role^*$, we select the
optimal centroids---we denote them by $\mu'$---among the profiles of the
vertices in the cluster. We know that the cost of $\role^*$ w.r.t. $\mu'$, say
$\tau$, is at most $2\sigma$. If $\tau \geq 8m$, then $\sigma \geq 4m$, the cost of $\role$, which
violates the optimality of $\role^*$. Consequently, $\tau < 8m = t$.

Note that $\mu'$ are all integral. We can safely assume that the copies of the
first part of the graph have the same role assignment. This immediately implies
that the profiles of $\role^*$ should match \emph{exactly} the centroids $\mu'$.

There are 6 groups with distinct degrees in the first part:
$\dg{u} = t\ell$ for $u \in A_1$,
$\dg{u} = 3t$ for $u \in A_2$,
$\dg{u} = 1 + 3t$ for $u \in A_3$,
$\dg{u} = 2s + 1$ for $u \in A_4$,
$\dg{u} = s + 1$ for $u \in A_5$,
$\dg{u} = 2$ for $u \in A_6 \cup A_7$.
Since the vertices from different groups have different degrees, $\role^*$
must be a refinement of this partition.
Moreover, $A_6$ connects to $A_2$ and $A_7$ connects to $A_3$.
Thus they cannot have the same role. 
This gives us 7 groups, and $\role^*$ must be a refinement of this partition.
Since we have only 7 roles, this must be the role assignment.

Let us now consider the second part of the graph.
Let $u \in B_1$.
The only integral centroid that matches $\dg{u}$ is $\mu'_1$, and the
remaining centroids differ in degree
by at least of $t$, consequently $\dist{u, \mu'_i} \geq t$ for $i \neq 3$.
This forces, $\role^*(u) = 1$. Similarly,
$\role^*(u) = 4$, for $u \in B_4$, 
$\role^*(u) = 5$, for $u \in B_5$,
$\role^*(u) = 6, 7$, for $u \in B_6 \cup B_7$.

Let $u \in B_2 \cup B_3$. If $\role^*(u) \neq 2, 3$, then each vertex in a fat edge will
introduce a cost of at least 1, when compared to the integral centroids $\mu'$.
Since there are $t$ of these vertices, we must have $\role^*(u) = 2, 3$.  Using
a similar argument, among the vertices in $B_2 \cup B_3$ adjacent (by a fat edge) to a vertex $v \in B_1$,
only one has the role $3$, the remaining adjacent vertices have the role $2$.

This shows that $\role^* = \role_{\mathcal{W}}$, where $\mathcal{W}$ are the sets corresponding
to the vertices with role $3$. We saw earlier that the cost is minimized when $\alpha = t +
\abs{\mathcal{W} \cap \mathcal{M}}$
is maximized.
\end{proof}


\iffalse
\emph{(i)} a vertex $o$ with $\dg{o} = 2s + 1$,
\emph{(ii)} a vertex in $K(s, s)$ with $\dg{o} = s + 1$,
\emph{(iii)} vertices $a \in A$ with $\dg{a} = t\ell$,
\emph{(iv)} vertex $b^*$ with $\dg{b^*} = 1 + 3t$,
\emph{(v)} the remaining vertices $b \in B \setminus \set{b^*}$ with $\dg{b} = 3t$, and finally
\emph{(vi)} the vertices in fat edges with degree of $2$.
Since the vertices from different groups have different degrees, $\role^*$
must be a refinement of this partition. Note that the fat edges can be split further into
two groups:
\emph{(vi.a)} vertices connecting to $b^*$, and
\emph{(vi.b)} vertices connecting to $B \setminus \set{b^*}$.
This gives us 7 groups, and $\role^*$ must be a refinement of this partition.
Since we have only 7 roles, this must be the role assignment.


Let us now consider the second part of the graph.
Let $a \in A'$.
The only integral centroid that matches $\dg{a}$ is $\mu'_3$, and the
remaining centroids differ in degree
by at least of $t$, consequently $\dist{a, \mu'_i} \geq t$ for $i \neq 3$.
This forces, $\role^*(a) = 3$. Similarly, each vertex $o \in O'$ has
$\role^*(o) = 1$, and the vertices in the bi-cliques have the role 2.
Each vertex $x$ in a fat edge must have $\role^*(x) = 6$ or $\role^*(x) = 7$.

Let $b \in B'$. If $\role^*(b) \neq 4, 5$, then each vertex in a fat edge will
introduce a cost of at least 1, when compared to the integral centroids $\mu'$.
Since there are $t$ of these vertices, we must have $\role^*(b) = 3, 4$.  Using
a similar argument, among the vertices in $B'$ adjacent to a vertex $a \in A'$,
only one has the role $4$, the remaining adjacent vertices have the role $5$.


The cost of $\role^*$ is now lower-bounded by the cost given in
Eq.~\ref{eq:cost}, and it is equal if and only if $\role^* = \role$.

Since the cost of $\role$ decreases as $\abs{D}$ increases, the optimal $\role$
will correspond to the optimal matching.
\end{proof}
\fi

To prove Proposition~\ref{proposition:np-fixed},
we need to introduce the following \np-complete problem.

\begin{problem}[\tuples]
Assume a universe $U$ and a set $\mathcal{S}$ of 5-tuples over $U$,
such that each $u \in U$ occurs in at least 3 tuples.
Is there a subset $B \subseteq U$ such that each exactly 2 entries in each $S \in \mathcal{S}$
belong to $B$.
\end{problem}

\begin{proposition}
\tuples is \np-complete.
\end{proposition}

\begin{proof}
\tuples is obviously in \np. We prove the hardness by reduction from \tsat.

Assume that we are given $m$ clauses $C_1, \ldots, C_m$ using $n$ variables, in
total. We can safely assume that each clause contains exactly 3 variables by
allowing the repetition of a variable.

We define the universe $U$ to contain $2n + 2 + 2m$ variables, which we will
denote by $t$, $f$, $x_1, \ldots, x_n$, $y_1, \ldots, y_n$,
$d_1, \ldots, d_m$, and $e_1, \ldots, e_m$.
To simplify notation,
we will identify $x_i$ with the $i$th positive literal and $y_i$ with the $i$th negative literal. 

We introduce the following clauses:
\emph{(1)} $(t, t, f, f, f)$,
\emph{(2)} $(x_i, \neg x_i, t, f, f)$, where $i = 1, \ldots, n$,
\emph{(3)}
$(\neg c_1, \neg c_2, \neg c_3, d_j, e_j)$, where $C_j = c_{1} \lor c_{2} \lor c_{3}$
is the $j$th clause.

To make sure that each variable occurs in at least 3 tuples, we copy each tuple
3 times.


Assume there is $B$ solving \tuples. Set the $i$th variable to be true if $x_i \in B$, and false otherwise.
At most $2$ variables $\neg c_i$ are in $B$, for $C_j = c_{1} \lor c_{2} \lor c_{3}$.
Assume that $\neq c_1 = \notin B$.
Since either $x_i \in B$ or $\neg x_i \in B$, this forces $c_1 \in B$, 
making $C_j$ satisfied.

Now assume that we can satisfy each clause. First set $B = \set{t}$. Add
$x_i$ into $B$ if the $i$th variable is true, otherwise insert $\neg x_i$.
Since a clause $C_j = c_{1} \lor c_{2} \lor c_{3}$ is satisfied, we have at most 2 entries in $B$ among $\neg c_i$. 
Insert $d_j$ and/or $e_j$ to $B$ so that $(\neg c_1, \neg c_2, \neg c_3, d_j, e_j)$ contains exactly 2 entries in $B$.
The resulting $B$ solves \tuples.
\end{proof}

\begin{proof}[of Proposition~\ref{proposition:np-fixed}]
The problem is in \np.  We will prove the hardness from \tuples.
Assume that we are given an instance $(U, \mathcal{S})$ of \tuples.
Let $n = \abs{U}$ and $m = \abs{\mathcal{S}}$.
Let $\ell_u$ be the total number of occurrences, counting multiplicities,
of $u$ in $\mathcal{S}$.

Define the graph as follows:
For each $u \in U$, add a cycle of length $\ell_u$. Note that the cycle is simple
since $\ell_u \geq 3$. Denote this cycle by $D_u$.

For each tuple $C_j$ add a vertex $w_j$, and connect it to 
a vertex in $D_u$, where $u \in C_j$. The connection can and should be done such
that each vertex in each cycle is connected to exactly one $w_j$.

Set the number of roles $k = 3$, and define the following centroids $\mu_1 = (2, 0, 1)$, $\mu_2 = (0, 2, 1)$,
and $\mu_3 = (2, 3, 0)$.

Let $r$ be the optimal role assignment for the defined centroids.
The cost of is 0 if and only if
\emph{(i)} every vertex in $v \in D_u$ has the same role, $\role(v) = 1, 2$,
\emph{(ii)} $\role(w_j) = 3$,
\emph{(iii)} for each $w_j$ there are two adjacent vertices with role 1
and three adjacent vertices with role 2.

Consequently, $B$ in \tuples corresponds to the vertices with role 1.
\end{proof}
